\vspace{-15mm}\fontsize{24pt}{10pt}\selectfont\textbf{A Radial Time Projection Chamber for $\alpha$ detection in CLAS at JLab}

### 2017-06-27 eg6-letters/NIM/EG6_RTPC_NIM.tex
Minor edits in NIM paper

Until recently, the Thomas Jefferson National Accelerator Facility, in 
Newport News, Virginia, USA, has provided high power electron beams of 
up to 6 GeV energy and 100$\%$ duty factor to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{CLASref}, located
in Hall-B, is based on a superconducting toroidal magnet and composed of 
several sub-detectors. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers for the tracking of charged 
       particles~\cite{DCref}.
 \item Superconducting toroidal magnet to bend the trajectories 
       of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Threshold Cherenkov Counters for electron identification at momenta 
       $<2.7$ GeV~\cite{CCref}.
 \item Scintillation Counters to identify charged hadrons by measuring their 
       time of flight~\cite{TOFref}.
 \item Electromagnetic Calorimeters (EC) for identification of electrons, 
       photons and neutrons~\cite{ECref}.
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a nucleon or nuclei) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005~\cite{FX}.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in RTPC. This 
setup was used during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC~\cite{BONUS-NIM} was developed for the BoNuS 
experiment at Jefferson Lab which took data in CLAS in 2005. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration strategy and results for 
the RTPC are discussed in section \ref{sec_calib}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230 mm diameter bore of the 
5 Tesla solenoid magnet.

@@

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam line axis; it 
      is a 284~mm long, 6~mm in diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at 4260~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower Voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

One of the important challenge in developing the radial TPC was how to obtain a 
good support structure for the GEM foils to allow a tractable 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each end of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the upstream end of the cylinder 
was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
parallel to the GEMs, 69 mm from the central axis.
Figure~\ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width. The shift between the rows permits the reduction of aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads are grouped to 16 channel pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and driv the 6~m long ribbon 
cable that connects to it.

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture~\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time and accounted for it in the drift 
speed used for the track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate them, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.20 and 1.27 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate directly the kinematic of the Helium nucleus. 
By comparing the alpha particle's calculated momentum and angle of recoil to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the exact chamber's conditions.

@@

The gain of each pad is defined as the ratio between the actual 
deposited energy and the registered ADC value. 
To do so we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}.

@@

@@

\section{Track Reconstruction}\label{sec_rec}
In order to reconstruct tracks we first select good hits. This means rejecting 
out-of-time hits as well as the hits linked to the electronic noise. The second step is 
to reconstruct the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit with a helix the chains that have a minimum of 
10 hits. We make the fit in two iterations, first with all the hits of the chain 
and the beam line, then, in the second iteration, 
the hits that are 5 mm or farther from the first fit are excluded.

@@

\subsection{Efficiency}
The previous discussions have shown that the two modules of the RTPC have 
slightly different behavior and give also different particle yield. However, 
this should not necessarily be linked to a different performance of the RTPC. 
Indeed, there is a complicated convolution of CLAS and the RTPC acceptances 
that can significantly affect the yields and kinematics of the detected 
particles. To clarify this question, we measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two modules of the detector. We observe that the left and 
the right modules have similar efficiencies except near the target windows.

@@

We reported on the construction, operation and calibration of a small RTPC 
designed to measure helium-4 nuclei in high rate environment. The operation
of the detector was successful and allowed to detect helium nuclei at a rate
of 3.1 kHz in coincidence with the detection of high energy electrons and 
photons in the CLAS spectrometer.

>>>>

Until recently, the Thomas Jefferson National Accelerator Facility, in 
Newport News, Virginia, USA, has provided high power electron beams of 
up to 6 GeV energy and 100$\%$ duty factor to three experimental Halls (A, B, C) 
simultaneously. The CLAS spectrometer~\cite{CLASref}, located
in Hall-B, is based on a superconducting toroidal magnet and composed of 
several sub-detectors. Figure~\ref{fig:CLAS} shows a three dimensional 
representation of the baseline CLAS spectrometer:
\begin{itemize}
 \item Three regions of Drift Chambers (DC) for the tracking of charged 
       particles~\cite{DCref}.
 \item Superconducting toroidal magnet to bend the trajectories 
       of charged particles, thus allowing momentum measurement with the DC tracking information.
 \item Threshold Cherenkov Counters for electron identification at momenta 
       $<2.7$ GeV~\cite{CCref}.
 \item Scintillation Counters to identify charged hadrons by measuring their 
       time of flight~\cite{TOFref}.
 \item Electromagnetic Calorimeters (EC) for identification of electrons, 
       photons and neutrons~\cite{ECref}.
\end{itemize}

@@

Measurement of the Deeply Virtual Compton Scattering (DVCS) process
($eH \rightarrow e' H' \gamma$, where $H$ is a nucleon or nuclei) necessitates an upgrade
of this setup.  With a 6 GeV electron beam, the majority of DVCS photons are produced
at very forward angles, where the acceptance of the EC is poor. To extend the detection range,
an inner calorimeter (IC) was built for the E01-113 experiment in 2005~\cite{FX}.
The IC is constructed from 424 lead-tungstate (PbWO$_{4}$) crystals, covering polar 
angles between 5$^{\circ}$ and 15$^{\circ}$~\cite{Hyon-suk}. To protect the 
CLAS detector and the IC from the large flux of low energy M{\o}ller 
electrons, a 5~T solenoid was placed around the target to shield the detectors. 
To detect recoiling $\alpha$ particles from the coherent DVCS on helium, a new 
radial time projection chamber (RTPC) was developed to track low energy nuclear 
fragments. The solenoid field was used to track and measure momentum of 
particles in RTPC. This 
setup was used during a three months experimental run~\cite{proposal1,proposal2}
in 2009 with a longitudinally polarized electron beam of 130~nA 
and energy of 6.064 GeV incident on a gaseous $^{4}$He target.

The original design of the RTPC~\cite{BONUS-NIM} was developed for the BoNuS 
experiment at Jefferson Lab which took data in CLAS in 2005. An enhanced design, 
used in the 2009 DVCS experiment, is described in section \ref{sec_design} 
of this paper. Significant improvements were made to the RTPC mechanical 
structure and fabrication technique that both increased the acceptance and 
reduced the amount of material in the path of the outgoing particles. The 
data acquisition system described in section \ref{sec_readout} was reconfigured 
to increase the event readout rate. The calibration methods 
are discussed in section \ref{sec_calib}. Finally, the overall performance
of the RTPC is described in section \ref{sec_perfor}.

@@

With a 6 GeV incident electron energy, the recoiling $^{4}$He nuclei from coherent 
DVCS have an average momentum around 300 MeV/c (12 MeV kinetic energy). Such low energy $\alpha$ 
particles are stopped very rapidly, so the RTPC was designed to be as close 
as possible to the target and fit inside the 230~mm diameter bore of the 
5 Tesla solenoid magnet.

@@

We detail here the different regions shown in Fig.~\ref{fig:RTPC_1_4} starting 
from the beam line towards larger radius:\\
\begin{itemize}
   \item The 6 atm helium-4 target extends along the beam line axis; it 
      is a 284~mm long, 6~mm diameter straw with a 27-$\mu$m Kapton wall.
   \item The first gas gap covers a radial range from 3 mm to 20 mm. It is 
      filled with $^{4}$He gas at 1~atm to minimize secondary interactions from
      M\o{}ller electrons scattered by the beam. This region is surrounded by 
      a 4 $\mu$m thick window made of grounded aluminized Mylar.
   \item The second gap region extends between 20~mm and 30~mm and is filled with the 
      gas mixture of 80$\%$ neon (Ne) and 20$\%$ dimethyl ether (DME). This region 
      is surrounded by a 4~$\mu$m thick window made of aluminized Mylar set at 4260~V 
      to serve as the cathode.
   \item The drift region is filled with the same Ne-DME gas mixture and extends 
      from the cathode to the first GEM, 60 mm away from the beam axis. The average 
      electric field in this region is perpendicular to the beam and about 550~V/cm.
   \item The electron amplification system is composed of three GEMs located at 
      radii of 60, 63 and 66 mm. In this configuration, the first GEM layer 
      serves as the anode and each subsequent GEM is set at a lower Voltage to
      obtain a strong ($\sim$1600 V/cm) electric field between the GEM foils. A 
      275~V bias is applied across each GEM for amplification.
   \item The readout board has an internal radius of 69~mm and collects charge
     from the GEMs. Pre-amplifiers are plugged directly on its outer side and
     transmit the signal to the data acquisition electronics.
\end{itemize}

@@

One of the important challenge in developing the radial TPC was to obtain a 
good support structure for the GEM foils to allow a tractable 
installation of the GEMs. At the same time, we wanted to 
keep the material budget small in the forward region where we detect other 
particles in subsequent detectors. We successfully realized these prerequisites 
by using fiber glass rings glued to each end of the GEM foils to form self 
supporting cylinders that could be installed independently in the RTPC after 
gluing and soldering operations. The rigidity of the GEM foils was enough
for the structure to be self-supporting and only the upstream end of the cylinder 
was fixed to the main mechanical structure. This design only left a 
light fiberglass ring in the downstream end, reducing to a minimum secondary 
interactions.

@@

The RTPC electron collection system has 3200 readout pads. These elements are
parallel to the GEMs, 69 mm from the central axis.
Figure~\ref{fig:PADs} shows a schematic drawing of the size and 
configuration of the pads. Each readout pad is 5 mm in length and 4.45 mm in 
width. The shift between the rows permits the reduction of aliasing. Each half of the 
RTPC has 40 rows and 40 columns of pads. The shaded region in Fig.~\ref{fig:PADs} 
shows how pads are grouped to 16 channel pre-amplifiers. The pre-amplifier boards, 
already employed in the BoNuS RTPC~\cite{BONUS-NIM}, serve the dual purpose of 
inverting the RTPC signals polarity -- from negative to positive -- to match the 
requirements of the subsequent readout system, and drive the 6~m long ribbon 
cable that connects to it.

@@

Due to the non perfect experimental conditions, in particular variations in the 
gas mixture~\footnote{It was noted during the experiment that leaks appeared
and the gas flow had to be increased in few occasions to compensate for this.}, 
the drift speed changes during the three months long experimental run.  
Figure~\ref{fig:Drift_run_number_1} shows the $T_{Max/2}$ values for 
individual runs (approximately 2 hours long). We observe a significant 
variation of the drift speed over time that need to be accounted for 
in track reconstruction.

@@

The drift path is the trajectory followed by the electrons released through 
ionization in the gas. Software exists to calculate them, in particular
MAGBOLTZ~\cite{MAGBOLTZ}, but it requires knowledge of the detector's 
geometry, gas mixture composition, and of course the electric and magnetic 
fields over the whole volume of the detector. We used such calculation as a 
first calibration, but, 
as can be seen with the drift speed, gas composition is far from stable in the 
chamber. Moreover, the 4 $\mu$m foil used as a cathode is easily deformed, such
that we expect the geometry accuracy to be of few millimeters, directly impacting 
our knowledge of the electric field. These problems, already 
encountered for the BoNuS RTPC calibration~\cite{BONUS-NIM}, motivated the 
acquisition of specific calibration runs. These were taken with a lower energy 
electron beam (1.204 and 1.269 GeV) to enhance the cross section of the elastic 
scattering ($e^{4}$He$\rightarrow e^{4}$He). In this process, the measurement of
the electron kinematic allows to calculate directly the kinematic of the Helium nucleus. 
By comparing the alpha particle's calculated momentum and angle of recoil to the 
measurement in the RTPC, we tuned the drift paths independently of our 
knowledge of the exact chamber's conditions.

@@

To calibrate the gains, we compare the experimental ADCs to the energy deposited 
for each pad individually in GEANT4 by similar simulated tracks (using the 
same elastic events as for the drift 
paths calibration). This requires a very good GEANT4 simulation 
including drift paths, but also the spread of the charges along the path
before reaching the pad, so that the simulated hits match the experimental 
ones. Moreover, the simulation has to match the DAQ features 
that can lead to cutting out hits. After setting the simulation properly, we 
compared simulation to experiment on an event by event basis as shown in 
Fig.~\ref{fig:EVENT_adc_tdc}. The gain for each pad is calculated 
as the ratio of the measured ADCs to the simulated deposited energy.  
Then, these gains are refined using correcting factors obtained from a 
sample of good tracks. For each track, we calculate the corrected 
energy deposit on a pad and compare it to the average
deposit recorded by the other pads. This last step was implemented to 
avoid any bias from this simulation based calibration method.
Final results are shown in Fig.~\ref{fig:dedx_p_exp_2nd}.

@@

\section{Track Reconstruction}

@@

\subsection{Track Fitting}\label{sec_rec}
We start with reconstructing the spacial origin of the hits using the extracted drift speed and drift path 
parameters. For each registered hit, we calculate a position of emission from 
the recorded time and the position of the recording pad. The third step 
is to create chains of hits. The maximum distance between two close adjacent 
hits has to be less than 10.5 mm to chain them, which roughly corresponds to 
neighbors and next to neighbors. Then, we fit with a helix the chains that have a minimum of 
10 hits. We make the fit in two iterations, first with all the hits of the chain 
and the beam line, then, in the second iteration, 
the hits that are 5 mm or farther from the first fit are excluded.

@@

\subsection{Efficiency}
The previous discussions have shown that the two modules of the RTPC have 
slightly different behavior and give also different particle yield. However, 
this should not necessarily be linked to different efficiencies within the RTPC. 
Indeed, there is a complicated convolution of CLAS and the RTPC acceptances 
that can significantly affect the yields and kinematics of the detected 
particles. To clarify this question, we measured the efficiency of the RTPC
using elastic scattering on $^4$He by comparing the inclusive yield, based
only on electron detection, to the exclusive elastic yield, where the Helium
recoil is also detected (see Fig.~\ref{fig:w}). We present in Fig.~\ref{fig:rtpc_eff}
the results for the two modules of the detector. We observe that the left and 
the right modules have similar efficiencies except near the target windows.

@@

We reported on the construction, operation and calibration of a small RTPC 
designed to measure helium-4 nuclei in high rate environment. The operation
of the detector was successful and allowed to detect helium nuclei with a 75\%
efficiency and a readout rate
of 3.1 kHz triggered by the detection of high energy electrons and 
photons in the CLAS spectrometer.
